Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2015-01-01 papers/wdmerger_I/paper.tex
Slight clarification of how SPH handles the KH instability.

Given the above, why is another approach using a different simulation code warranted? First and foremost, reproducibility of the results across simulation codes and algorithms is important for gauging confidence in this result. With only a couple of exceptions, most of the existing results that study the viability of the DD progenitor as an explanation for Type Ia supernovae (that is, including a temperature-dependent equation of state and nuclear reactions) have used the SPH method. SPH codes have a number of features which do aid them in the study of these systems, such as excellent conservation of angular momentum. However, the question of whether a prompt detonation in a merger actually happens depends in detail on the nature of the gas at the interface between the two stars, which is at much lower density than the rest of the stellar material. The SPH codes use uniform mass particles, so their effective resolution is \textit{lowest} at the stellar surface. In contrast, a grid-based code with adaptive mesh refinement can zoom in on the regions where hotspots will develop, while also maintaining high enough resolution in the high-density regions to adequately capture the large-scale mass transfer dynamics. However, there are also outstanding questions of convergence in SPH (e.g. \cite{zhu_SPH:2014}) and whether the method correctly captures fluid instabilities. This is an important question for white dwarf mergers because of the likely importance small-scale instabilities will have on the evolution of the low-density gas at the primary's surface. The pioneering work of \cite{agertz:2007} compared grid and SPH codes and found some important differences. Most relevant for this discussion is that the SPH codes could not adequately handle mixing from the Kelvin-Helmholtz instability in the test they propose. \cite{price:2008} attributed this to the fact that the standard SPH evolution equations do not have a mechanism for capturing discontinuities in internal energy, and showed that the addition of an artificial thermal conductivity can dramatically improve the ability of the SPH codes to exhibit this instability. There have since been a number of papers discussing this issue, but to our knowledge none of these improvements have yet been incorporated into an SPH model of a WD merger. Another reason to be skeptical is that the detonation itself was inserted manually in \cite{pakmor:2010} based on small-scale detonation simulations. A self-consistent evolution of a thermonuclear detonation, if it existed, would significantly build confidence in the progenitor model.

>>>>

Given the above, why is another approach using a different simulation code warranted? First and foremost, reproducibility of the results across simulation codes and algorithms is important for gauging confidence in this result. With only a couple of exceptions, most of the existing results that study the viability of the DD progenitor as an explanation for Type Ia supernovae (that is, including a temperature-dependent equation of state and nuclear reactions) have used the SPH method. SPH codes have a number of features which do aid them in the study of these systems, such as excellent conservation of angular momentum. However, the question of whether a prompt detonation in a merger actually happens depends in detail on the nature of the gas at the interface between the two stars, which is at much lower density than the rest of the stellar material. The SPH codes use uniform mass particles, so their effective resolution is \textit{lowest} at the stellar surface. In contrast, a grid-based code with adaptive mesh refinement can zoom in on the regions where hotspots will develop, while also maintaining high enough resolution in the high-density regions to adequately capture the large-scale mass transfer dynamics. However, there are also outstanding questions of convergence in SPH (e.g. \cite{zhu_SPH:2014}) and whether the method correctly captures fluid instabilities. This is an important question for white dwarf mergers because of the likely importance small-scale instabilities will have on the evolution of the low-density gas at the primary's surface. The pioneering work of \cite{agertz:2007} compared grid and SPH codes and found some important differences. Most relevant for this discussion is that the SPH codes could not adequately handle mixing from the Kelvin-Helmholtz instability in the test they propose. As pointed out by \cite{price:2008}, this is not a result of SPH being inherently unable to model this instability, but instead it is attributed to the fact that the standard SPH evolution equations do not have a mechanism for capturing discontinuities in internal energy. \citeauthor{price:2008} showed that the addition of an artificial thermal conductivity can dramatically improve the ability of the SPH codes to exhibit this instability. There have since been a number of other papers discussing this issue, but to our knowledge none of these improvements have yet been incorporated into an SPH model of a WD merger. Another reason to be skeptical is that the detonation itself was inserted manually in \cite{pakmor:2010} based on small-scale detonation simulations. A self-consistent evolution of a thermonuclear detonation, if it existed, would significantly build confidence in the progenitor model.
